---
editor_options: 
  chunk_output_type: console
---

# Data manipulation with `dplyr`

```{r load_lib}
# load the tidyverse
library(tidyverse)
```

## Introduction
### Foreword on `dplyr`
`dplyr` is tasked with performing all sorts of transformations on a dataset.

The structure of `dplyr` revolves around a set of functions, the so-called **verbs**, that share a common syntax and logic, and are meant to work with one another in chained operations. Chained operations are performed with the pipe operator (`%>%`), that will be introduced in section 3.2.2.

The basic syntax is `verb(data, variable)`, where `data` is a data frame and `variable` is the name of one or more columns containing a set of values for each observation.

There are 5 main verbs, which names already hint at what they do: `rename()`, `select()`, `filter()`, `mutate()`, and `summarise()`.
I'm going to introduce each of them (and a couple more) through the following sections.

### Example data

Through this tutorial, we will be using mammal trait data from the [Phylacine](https://megapast2future.github.io/PHYLACINE_1.2/) database.
Let's have a peek at what it contains.

```{r read_data, message=FALSE}
phylacine <- read_csv("data/phylacine_traits.csv")
phylacine
```

`readr` automatically loads the data in a `tibble`, as we have seen in chapter
1 and 2. Calling the tibble gives a nice preview of what it contains. We have 
data for 5,831 mammal species, and the variables contain information on taxonomy,
(broad) habitat, mass, IUCN status, and diet.

If you remember Section 1.2 on tidy data, you may see that this data isn't 
exactly tidy. In fact, some columns are in wide (and messy) format, like the 
"habitat" (terrestrial, marine, etc.) and diet columns. 

`dplyr` actually does not require your data to be strictly tidy. If you feel that your
data satisfies the definition "one observation per row, one variable per column",
that's probably good enough. 

I use a `tibble` here, but `dplyr` works equally well on base data frames. In
fact, `dplyr` is built for `data.frame` objects, and tibbles are data frames.
Therefore, tibbles are mortal.

## Working with existing variables

### Renaming variables with `rename()`

The variable names in the phylacine dataset are descriptive, but quite unpractical. Typing 
`Binomial.1.2.` is cumbersome and subject to typos (in fact, I just made one). 
`binomial` would be much simpler to use.

Changing names is straightforward with `rename()`.

```{r rename}
rename(
  .data = phylacine, 
  "binomial" = Binomial.1.2,
)


```

The first argument is always `.data`, the data table you want to apply change to. 
Note how columns are referred to. Once the data table as been passed as an 
argument, there is no need to refer to it directly anymore, `dplyr` understands 
that you're dealing with variables inside that data frame. So drop that 
`data$var`, `data[, "var"]`, and forget the very existence of `attach()` / 
`detach()`. 

You can refer to variables names either with strings or directly as objects, 
whether you're reading or creating them:

```{r rename2, eval=TRUE}
rename(
  phylacine,
  # this works
  binomial = Binomial.1.2
)
rename(
  phylacine,
  # this works too!
  binomial = "Binomial.1.2"
)
rename(
  phylacine,
  # guess what
  "binomial" = "Binomial.1.2"
)
```

I have applied similar changes to all variables in the dataset. Here is what the
new names look like:

```{r rename3, include=FALSE, echo=TRUE}
phylacine <- rename(
  .data = phylacine, 
  "binomial" = Binomial.1.2,
  "order" = Order.1.2,
  "family" = Family.1.2,
  "genus" = Genus.1.2,
  "species" = Species.1.2,
  "terrestrial" = Terrestrial,
  "marine" = Marine,
  "freshwater" = Freshwater,
  "aerial" = Aerial,
  "life_habit_method" = Life.Habit.Method,
  "life_habit_source" = Life.Habit.Source,
  "mass_g" = Mass.g,
  "mass_method" = Mass.Method,
  "mass_source" = Mass.Source,
  "mass_comparison" = Mass.Comparison,
  "mass_comparison_source" = Mass.Comparison.Source,
  "island_endemicity" = Island.Endemicity,
  "iucn_status" = IUCN.Status.1.2, # not even for acronyms
  "added_iucn_status" =  Added.IUCN.Status.1.2,
  "diet_plant" = Diet.Plant,
  "diet_vertebrate" = Diet.Vertebrate,
  "diet_invertebrate" = Diet.Invertebrate,
  "diet_method" = Diet.Method,
  "diet_source" = Diet.Source
)
phylacine
```

### The pipe operator `%>%`

If you have already come across pieces of code using the tidyverse, chances are 
that you have seen this odd symbol. While the pipe is not strictly-speaking a
part of the tidyverse (it comes from its own package, `magrittr`), it is 
imported along with each package and widely used in conjunction with its 
functions.
What does it do? Consider the following example with `rename()`:

```{r the_pipe}
phylacine2 <- readr::read_csv("data/phylacine_traits.csv")
# regular syntax
rename(
  phylacine2,
  "binomial" = "Binomial.1.2"
)
# alternative syntax with the pipe operator
phylacine2 %>% 
  rename(
    "binomial" = "Binomial.1.2"
  )
```

Got it? The pipe takes the object on its left-side and silently feeds it to the
*first* argument of the function on its right-side. It could be read as "take x,
then do...".
The reason for using the pipe is because it makes code syntax closer to 
the syntax of a sentence, and therefore, easier and faster for your brain to
process (and write!) the code. In particular, the pipe enables easy chains of 
operations, where you apply something to an object, then apply something else to
the outcome, and so on...
Through the later sections, you will see some examples of chained operations with `dplyr` functions, but
for that I first need to introduce a couple more verbs.

Using the pipe can be quite unsettling at first, because you are not used to
think in this way. But if you push a bit for it, I promise it will make things
a lot easier (and it's quite addictive!). To avoid typing the tedious symbols,
`magrittr` installs a shortcut for you in RStudio. Use `Ctrl + Shift + M` on 
Windows, and `Cmd + Shift + M` on MacOS.

Finally I should emphasize that the use of the pipe isn't limited to the 
tidyverse, but extends to almost all R functions. Remember that by default
the piped value is always matched to the first argument of the following 
function

```{r pipe_examples}
5 %>% rep(3)
"meow" %>% cat()
```

If you need to pass the left-hand side to an argument other than the first,
you can use the dot place-holder `.`.

```{r dot}
"meow" %>% cat("cats", "go", .)
```

Because of its syntax, most base R operators are not compatible with the pipe (but this is very rarely needed). If needed, `magrittr` introduces alternative functions for operators.

Subsetting operators can be piped, with the dot place-holder.

```{r pipe_limits}
# 5 %>% * 3 # no, won't work
# 5 %>% .* 3 # neither
5 %>% magrittr::multiply_by(3) # yes

# subsetting
list("monkey see", "monkey_do") %>% .[[2]]
phylacine %>% .$binomial %>% head()
```

Because subsetting in this way is particularly hideous, `dplyr`
delivers a function to extract values from a single variable. In only works on tables, though.

```{r pull}
phylacine %>% pull(binomial) %>% head()
```

### Select variables with `select()`

To extract a set of variables (i.e. columns), use the conveniently-named 
`select()`. The basic syntax is the same as `rename()`: pass your data as the 
first argument, then call the variables to select, quoted or not.

```{r select}
# Single variable
phylacine %>% select(binomial)
# A set of variables
phylacine %>% select(genus, "species", mass_g)
# A range of contiguous variables
phylacine %>% select(family:terrestrial)
```

You can select by variable numbers. This is not recommended, as prone to 
errors, especially if you change the variable order.

```{r select_nb}
phylacine %>% select(2)
```

`select()` can also be used to *exclude* variables:

```{r exclude}
phylacine %>% select(-binomial)
phylacine %>% select(-(binomial:species))
```

`select()` and `rename()` are pretty similar, and in fact, `select()` can also 
rename variables along the way:

```{r select_rename}
phylacine %>% select("linnaeus" = binomial)
```

And you can mix all of that at once:

```{r select_mix}
phylacine %>% select(
  "fam" = family,
  genus:freshwater,
  -terrestrial
)
```

### Select variables with helpers

The Rstudio team just released `dplyr 1.0.0`, and along with it, some nice 
helper functions to ease the selection of a set of variables. I give three 
examples here, and encourage you to look at the documentation (`?select()`) to
find out more.

```{r select_help}
phylacine %>% select(where(is.numeric))
phylacine %>% select(contains("mass") | contains("diet"))

habitats <- c("terrestrial", "marine", "arboreal", "fossorial", "freshwater")
phylacine %>% select(any_of(habitats))

```

### Rearranging variable order with `relocate()`

The order of variables seldom matters in `dplyr`, but due to popular demand, 
`dplyr` now has a dedicated verb to rearrange the order of variables. The syntax
is identical to `rename()`, `select()`.

```{r relocate}
phylacine %>% relocate(mass_g, .before = binomial)
phylacine %>% relocate(starts_with("diet"), .after = species)
```


## Working with observations

### Ordering rows by value - `arrange()`

`arrange()` sorts rows in the data by **ascending** value for a given variable.
Use the wrapper `desc()` to sort by descending values instead.

```{r}
# Smallest mammals
phylacine %>% 
  arrange(mass_g) %>% 
  select(binomial, mass_g)

# Largest mammals
phylacine %>% 
  arrange(desc(mass_g)) %>% 
  select(binomial, mass_g)

# Extra variables are used to sort ties in the first variable
phylacine %>% 
  arrange(mass_g, desc(binomial)) %>% 
  select(binomial, mass_g)
```

*Important*: `NA` values, if present, are always ordered at the end!

### Subset rows by position - `slice()`

Use `slice()` and its variants to extract particular rows.

```{r slice}
phylacine %>% slice(3) # third row
phylacine %>% slice(5, 1, 2) # fifth, first and second row
phylacine %>% slice(rep(3, 2)) # duplicate the third row
phylacine %>% slice(-c(2:5830)) # exclude all but first and last row

phylacine %>% slice_tail(n = 3) # last three rows
phylacine %>% slice_max(mass_g) # largest mammal
```

You can also sample random rows in the data:

```{r slice_sample}
phylacine %>% slice_sample() # a random row

# bootstrap
phylacine %>% slice_sample(n = 1000, replace = TRUE)

```

### Subsetting rows by value with `filter()`

`filter()` does a similar job as `slice()`, but extract rows that satisfy a set
of conditions. The conditions are supplied much the same way as you would do 
for an `if` statement.

Along with `mutate()` (next section), this is probably the function you are going to use 
the most.

For example, I might want to extract mammals above a given mass:

```{r megafauna}
# megafauna
phylacine %>% 
  filter(mass_g > 1e5) %>% # 100 kg
  select(binomial, mass_g)
# non-extinct megafauna
phylacine %>% 
  filter(mass_g > 1e5, iucn_status != "EP") %>% 
  select(binomial, mass_g, iucn_status)


```

Are there any flying mammals that aren't bats?

```{r no_bats}
phylacine %>% 
  filter(aerial == 1, order != "Chiroptera") 
# no :(
```

Are humans included in the table?

```{r human_after_all}
phylacine %>% filter(binomial == "Homo_sapiens")
```

`filter()` can be used to deal with NAs:

```{r filter_na}
phylacine %>% 
  filter(!is.na(mass_comparison)) 
```

Tip: `dplyr` introduces the useful function `between()` that does exactly what  the name implies

```{r between}
between(1:5, 2, 4)

# Mesofauna
phylacine %>% 
  filter(mass_g > 1e3, mass_g < 1e5) %>% 
  select(binomial, mass_g)

# same thing
phylacine %>% 
  filter(mass_g %>% between(1e3, 1e5)) %>% 
  select(binomial, mass_g)

```

Note that you can pipe operations inside function arguments as in the last line 
above (arguments are expressions, after all!).

## Making new variables

### Create new variables with `mutate()`

Very often in data analysis, you will want to create new variables, or edit 
existing ones. This is done easily through `mutate()`. For example, consider 
the diet data:

```{r}
diet <- phylacine %>% 
  select(
    binomial,
    diet_plant,
    diet_vertebrate,
    diet_invertebrate
  )
diet
```

These three variables show the percentage of each category of food that make the
diet of that species. They should sum to 100, unless the authors made a typo or
other entry error. To assert this, I'm going to create a new variable, 
`total_diet`.

```{r total_diet}
diet <- diet %>% mutate(
  "total_diet" = diet_vertebrate + diet_invertebrate + diet_plant
)
diet

all(diet$total_diet == 100)
# cool and good
```

`mutate()` adds a variable to the table, and keeps all other variables. 
Sometimes you may want to just keep the new variable, and drop the other ones.
That's the job of `mutate()`'s twin sibling, `transmute()`. For example,
I want to combine `diet_invertebrate` and `diet_vertebrate` together:

```{r diet_animal}
diet %>% 
  transmute(
    "diet_animal" = diet_invertebrate + diet_vertebrate
  )
```

You may want to keep some variables and drop others. You could pipe `mutate()` 
and `select()` to do so, or you could just pass the variables to keep to
`transmute()`.

```{r both_diets}
diet %>% 
  transmute(
    "diet_animal" = diet_invertebrate + diet_vertebrate,
    diet_plant
  )
```

You can also refer to variables you're creating to derive new variables from 
them as part of the same operation, this is not an issue.

```{r total_diet2}
diet %>% 
  transmute(
    "diet_animal" = diet_invertebrate + diet_vertebrate,
    diet_plant,
    "total_diet" = diet_animal + diet_plant
  )
```

Sometimes, you may need to perform an operation based on the row number 
(I don't have a good example in mind).
To add the row number as a variable, you could do:

```{r row_nb}
phylacine %>% 
  transmute(
    binomial,
    "row_nb" = 1:nrow(phylacine)
  )
```

but `tibble` has a built-in function to do just that:

```{r row_index}
phylacine %>% 
  select(binomial) %>% 
  tibble::rownames_to_column(var = "row_nb")
```

### Summarise observations with `summarise()`

`mutate()` applies operations to all observations in a table. By contrast, 
`summarise()` applies operations to *groups* of observations, and returns, er, summaries. The default grouping unit is the entire table:

```{r summarise}
phylacine %>% 
  summarise(
    "nb_species" = n(), # counts observations
    "nb_terrestrial" = sum(terrestrial),
    "nb_marine" = sum(marine),
    "nb_freshwater" = sum(freshwater),
    "nb_aerial" = sum(aerial),
    "mean_mass_g" = mean(mass_g)
  )
```

Above you can see that bats account for a large portion of mammal species 
diversity (`nb_aerial`). How much exactly? Just as with `mutate()`, you can 
perform operations on the variables you just created, in the same statement:

```{r prop_bats}
phylacine %>% 
  summarise(
    "nb_species" = n(),
    "nb_aerial" = sum(aerial), # bats
    "prop_aerial" = nb_aerial / nb_species
  )
```

One fifth!
If the british spelling bothers you, `summarize()` exists and is strictly 
equivalent.

Here's a simple trick with logical (TRUE / FALSE) variables. Their sum is the 
count of observations that evaluate to `TRUE` (because `TRUE` is taken as 1 and 
`FALSE` as 0) and their mean is the proportion of `TRUE` observations. This can 
be exploited to count the number of observations that satisfy a condition:

```{r sum_logi}
phylacine %>% 
  summarise(
    "nb_species" = n(),
    "nb_megafauna" = sum(mass_g > 100000),
    "p_megafauna" = mean(mass_g > 100000)
  )

```

There are more summaries that just means and counts (see `?summarise()` for some helpful functions). In fact, summarise can use any function or expression that evaluates to a single value or a *vector* of values. This includes base R `max()`, `quantiles`, etc.

`mutate()` and `transmute()` can compute summaries as well, but they will return the summary once for each observation, in a new column.

```{r mutate_summary}
phylacine %>% 
  mutate("nb_species" = n()) %>%
  select(binomial, nb_species)
```

### Grouping observations by variables

In most cases you don't want to run summary operations on the entire set of observations, but instead on observations that share a common value, i.e. groups.
For example, I want to run the summary displayed above, but for each Order of mammals. 

`distinct()` extracts all the unique values of a variable

```{r distinct_orders}
phylacine %>% distinct(order)
```

I could work my way with what we have already seen, filtering observations (`filter(order == "Rodentia")`) and then pipeing the output to `summarise()`, and do it again for each Order. But that would be tedious.

Instead, I can use `group_by()` to pool observations by `order`.

```{r group_by_order}
phylacine %>% 
  group_by(order)
```

At first glance, nothing has changed, apart from an extra line of information in the output that tells me the observations have been grouped. But now here's what happen if I run the same `summarise()` statement on an ungrouped and a grouped table

```{r summarise_order}
phylacine %>% 
  summarise(
    "n_species" = n(),
    "mean_mass_g" = mean(mass_g)
  )

phylacine %>% 
  group_by(order) %>% 
  summarise(
    "n_species" = n(),
    "mean_mass_g" = mean(mass_g)
  )
```

I get one value for each group. 

Observations can be grouped by multiple variables, which will output a summary for every unique combination of groups.

```{r multiple_groups}
phylacine %>% 
  group_by(order, iucn_status) %>% 
  summarise(
    "n_species" = n()
  )
```

Whenever you call `summarise()`, the last level of grouping is dropped. Note how in the output table above, observations are still grouped by order, and no longer by IUCN status. If I summarise observations again:

```{r drop_grouping}
phylacine %>% 
  group_by(order, iucn_status) %>% 
  summarise(
    "n_species" = n()
  ) %>% 
  summarise(
    "n_species_2" = n()
  )
```

I get the summary across orders, and the table is no longer grouped at all. This is useful to consider if you need to work on summaries across different levels of the data. 

For example, I would like to know how the species in each order are distributed between the different levels of threat in the IUCN classification. To get these proportions, I need to first get the count of each number of species in a level of threat inside an order, and divide that by the number of species in that order.

```{r iucn_prop}
iucn_props <- phylacine %>% 
  group_by(order, iucn_status) %>% 
  summarise("n_order_iucn" = n()) %>% 
  # grouping by iucn_status silently dropped
  mutate(
    "n_order" = sum(n_order_iucn),
    "p_iucn" = n_order_iucn / n_order
  )
iucn_props

iucn_props %>% 
  filter(order == "Carnivora")
```

10.2% of Carnivores are Endangered ("EN").

### Grouped data and other `dplyr` verbs

Grouping does not only affect the behaviour of `summarise`, but under circumstances, other verbs can (and will!) perform operations by groups.

```{r grouped_filter}
# Species with a higher mass than the mammal mean
phylacine %>% 
  select("binomial", "mass_g") %>% 
  filter(mass_g > mean(mass_g, na.rm = TRUE))

# Species with a higher mass than the mean in their order
phylacine %>% 
  group_by(order) %>% 
  select("binomial", "mass_g") %>% 
  filter(mass_g > mean(mass_g, na.rm = TRUE))

```

```{r grouped_arrange}
# Largest mammal
phylacine %>% 
  select(binomial, mass_g) %>% 
  slice_max(mass_g) 
# Largest species in each order
phylacine %>% 
  group_by(order) %>% 
  select(binomial, mass_g) %>% 
  slice_max(mass_g) 
```

To avoid grouped operations, you can simply drop grouping with `ungroup()`.

## Working with multiple tables

### Binding tables

`dplyr` introduces `bind_rows()` and `bind_cols()`, which are equivalent to base
R `rbind()` and `cbind()`, with a few extra feature. They are faster, and can 
bind many tables at once, and bind data frames with vectors or lists. 

`bind_rows()` has an option to pass a variable specifying which dataset each 
observation originates from.

```{r id_bind}
porpoises <- phylacine %>% 
  filter(family == "Phocoenidae") %>% 
  select(binomial, iucn_status)
echidnas <- phylacine %>% 
  filter(family == "Tachyglossidae") %>% 
  select(binomial, iucn_status)

bind_rows(
  "porpoise" = porpoises, 
  "echidna" = echidnas, 
  .id = "kind"
)
```

### Combining variables of two tables with mutating joins

Mutating joins are tailored to combine tables that share a set of observations
but have different variables. 

As an example, let's split the `phylacine` dataset in two smaller datasets, 
one containing information on diet and one on the dominant habitat.

```{r split_phylacine}
diet <- phylacine %>% 
  select(binomial, diet_plant:diet_invertebrate) %>% 
  slice(1:5)
diet

life_habit <- phylacine %>% select(binomial, terrestrial:aerial) %>% 
  slice(1:3, 6:7)
life_habit

```

The two datasets each contain 5 species, the first three are shared, and the 
two last differ between the two.

```{r intersect}
intersect(diet$binomial, life_habit$binomial)
setdiff(diet$binomial, life_habit$binomial)
```

To use mutate-joins, both tables need to have a **key**, a variable that 
identifies each observation. Here, that would be `binomial`, the sepcies names.
If your table doesn't have such a key and the rows between the tables match 
one another, remember you can create a row number variable easily with 
`tibble::column_to_rownames()`.

```{r inner_join}
inner_join(diet, life_habit, by = "binomial")
```

`inner_join` combined the variables, and dropped the observations that weren't 
matched between the two tables. There are three other variations of mutating 
joins, differing in what they do with unmatching variables.

```{r outer_joins}
left_join(diet, life_habit, by = "binomial")
right_join(diet, life_habit, by = "binomial")
full_join(diet, life_habit, by = "binomial")

semi_join(diet, life_habit, by  = "binomial")
anti_join(diet, life_habit, by  = "binomial")
```

### Filtering matching observations between two tables wiht filtering joins

So-called filtering joins return row from the first table that are matched 
(or not, for `anti_join()`) in the second.

```{r filter_joins}
semi_join(diet, life_habit, by  = "binomial")
anti_join(diet, life_habit, by  = "binomial")
```




