---
editor_options: 
  chunk_output_type: console
---

# Data manipulation with `dplyr`

```{r load_lib}
# load the tidyverse
library(tidyverse)
```

## Introduction

Reminders from last weeks: pipe operator, tidy tables, ggplot

Why dplyr ? dplyr vs base R

## Example data of the day

Through this tutorial, we will be using mammal trait data from the [Phylacine](https://megapast2future.github.io/PHYLACINE_1.2/) database. The dataset contains information on mass, diet, life habit, etc, for more than all living species of mammals. Let's have a look.

```{r read_data, message=FALSE}
phylacine <- readr::read_csv("data/phylacine_traits.csv")
phylacine
```

Note the friendly output given by the `tibble` (as opposed to a `data.frame`). `readr` automatically stores the content it reads in a `tibble`, tidyverse oblige. You should know however that `dplyr` doesn't require your data to be in a tibble, a regular `data.frame` will work just as fine.

Most of the `dplyr` verbs covered in the next sections assume your data is *tidy*: wide format, variables as column, 1 observation per row. Not that tehy won't work if your data isn't tidy, but the results could be very different from what I'm going to show here. Fortunately, the phylacine trait dataset appears to be tidy: there is one unique entry for each species.

The first operation I'm going to run on this table is changing the names with
`rename()`. Some people prefer their tea without sugar, and I [prefer](https://style.tidyverse.org/syntax.html#object-names) my 
variable names without uppercase characters, dots or (if possible) numbers. 
This will give me the opportunity to introduce the trivial syntax
of `dplyr` verbs.

```{r rename}
phylacine <- phylacine %>% 
  dplyr::rename(
    "binomial" = Binomial.1.2,
    "order" = Order.1.2,
    "family" = Family.1.2,
    "genus" = Genus.1.2,
    "species" = Species.1.2,
    "terrestrial" = Terrestrial,
    "marine" = Marine,
    "freshwater" = Freshwater,
    "aerial" = Aerial,
    "life_habit_method" = Life.Habit.Method,
    "life_habit_source" = Life.Habit.Source,
    "mass_g" = Mass.g,
    "mass_method" = Mass.Method,
    "mass_source" = Mass.Source,
    "mass_comparison" = Mass.Comparison,
    "mass_comparison_source" = Mass.Comparison.Source,
    "island_endemicity" = Island.Endemicity,
    "iucn_status" = IUCN.Status.1.2, # not even for acronyms
    "added_iucn_status" =  Added.IUCN.Status.1.2,
    "diet_plant" = Diet.Plant,
    "diet_vertebrate" = Diet.Vertebrate,
    "diet_invertebrate" = Diet.Invertebrate,
    "diet_method" = Diet.Method,
    "diet_source" = Diet.Source
  )
```

For convenience, I'm going to use the pipe operator (`%>%`) that we've seen 
before, through this chapter. All `dplyr` functions are built to work with the 
pipe (i.e, their firstargument is always `data`), but again, this is not 
compulsory. I could do

```{r no_pipe, eval=FALSE}
phylacine <- dplyr::rename(
  data = phylacine,
  "binomial" = Binomial.1.2,
  # ...
)
```
Note how columns are referred to. Once the data as been passed as an argument,
no need to refer to it anymore, `dplyr` understands that you're dealing with
variables inside that data frame. So drop that `data$var`, `data[, "var"]`,
and, if you've read *The R book*, forget the very existence of `attach()`. 

Finally, I should mention that you can refer to variables names either with strings or directly as objects, whether you're reading or creating them:

```{r rename2, eval=FALSE}
phylacine2 <- readr::read_csv("data/phylacine_traits.csv")

phylacine2 %>% 
  dplyr::rename(
    # this works
    binomial = Binomial.1.2
  )
phylacine2 %>% 
  dplyr::rename(
    # this works too!
    binomial = "Binomial.1.2"
  )
phylacine2 %>% 
  dplyr::rename(
    # guess what
    "binomial" = "Binomial.1.2"
  )
```

## Select variables with `select()`

To extract a set of variables (i.e. columns), use the conveniently-named `select()`.

```{r select}
phylacine_subset <- phylacine %>% 
  dplyr::select(
    binomial, 
    order, 
    terrestrial, 
    marine, 
    freshwater, 
    aerial
  )
phylacine_subset
# Single variable
phylacine %>% dplyr::select(family)
# A set of variables
phylacine %>% dplyr::select(genus, species, mass_g)
# A range of variables
phylacine %>% dplyr::select(family:terrestrial)
```

`select()` can also *exclude* variables:

```{r exclude}
phylacine %>% dplyr::select(-binomial)
phylacine %>% dplyr::select(-(binomial:species))
```

`select()` and `rename()` are pretty similar, and in fact, `select()` can also rename variables along the way:

```{r select_rename}
phylacine %>% dplyr::select("fam" = family)
```

And you can mix all of that at once:

```{r select_mix}
phylacine %>% dplyr::select(
  "fam" = family,
  genus:freshwater,
  -terrestrial
)
```

## Select observations with `filter()`

Conditional selection of observations is performed through `filter()`. This i sarguably the most useful function in the entire package. The syntax uses conditions involving the variables, just as you would use for `if` statements or `while` loops.

For example, I might want to extract mammals that are above a certain mass:

```{r megafauna}
# megafauna
phylacine %>% 
  dplyr::filter(mass_g > 1e5) # 100 kg
# living megafauna
phylacine %>% 
  dplyr::filter(mass_g > 1e5, iucn_status != "EP")

# Are there any flying mammals besides bats?
phylacine %>% 
  dplyr::filter(aerial == 1, order != "Chiroptera") 
# no :(

# That one species
phylacine %>% dplyr::filter(binomial == "Homo_sapiens")

```

Tip: `dplyr` introduces the useful function `between()` that does exactly what the name implies

```{r between}
between(1:5, 2, 4)

# Mesofauna
phylacine %>% dplyr::filter(mass_g > 1e3, mass_g < 1e5) # base
phylacine %>% dplyr::filter(mass_g %>% between(1e3, 1e5)) # dplyr

```

If instead of selecting observations based on conditions, you want to get the nth row, use `slice()`

```{r slice}
phylacine %>% dplyr::slice(1:3)
# can also be used to exclude rows
phylacine %>% 
  dplyr::slice(1:3) %>% 
  dplyr::slice(-2) 
```

You may also want the row number to be included as a variable, to be used as part of some condition or calculation. `tibble` is there for you:

```{r row_index}
phylacine %>% tibble::rownames_to_column()
```


## Create new variables with `mutate()`
can also edit existing ones

drop existing variables with `transmute()`

## Grouped results with `group_by()` and `summarise()`

## Scoped variables

```{r eval=FALSE}
data(mtcars)
mtcars %>% select_all(toupper)

is_whole <- function(x) all(floor(x) == x)
mtcars %>% select_if() # select integers only

mtcars %>% select_at(vars(-contains("ar")))
mtcars %>% select_at(vars(-contains("ar"), starts_with("c")))

```


## More !
dolla sign x point operator
variables values -> dplyr::distinct() eq. to base::unique()
sample_n()
sample_frac()
first, last, nth
